# (PART) Getting started {-}

# Introduction

**Learning objectives:**

1. Explain the **book club format.**
2. Explain **what** an **R package** is.
3. Describe **why** we should **develop packages.**
4. Discuss the philosophy behind the use of the **`devtools` and `usethis` packages** in **package development.**

## What is an R package?

* R packages can be defined based on their utility:
   * A means to bundle code, data, documentation, and tests that can easily be shared with others. 
   * A package opens up the use of several workflow packages/tools.
   
* R packages can be defined by their structure:
   * Discussed throughout the book club
   * At its core, packages are defined by the conventions of how files and directories are organized: `\R`; `DESCRIPTION` file; `\tests`; `\data`; etc. 
   * [Packaging Data Analytical Work Reproducibly Using R (and Friends)](https://www.tandfonline.com/doi/abs/10.1080/00031305.2017.1375986)

## Installing and using packages

* Packages are available via [Comprehensive R Archive Network (CRAN)](https://cran.r-project.org/index.html)

* Packages are also available via different public repos
   - [GitHub is the most popular](https://github.com/).

### Installing packages

```{r 01-install-packages, eval=FALSE}
install.packages("x")
library("x")

# Getting help
package?x
help(package = "x")
```

## Why develop packages?

* Allows us to more easily share and use code amongst the team and other collaborators.
* Saves yourself and your team's time.
* Opens up the use of several tools to more efficiently write code.
* Defines a formal organizational structure of where to put things. 
* Allows us to develop more robust, rigorous, reproducible analyses.
* Helps us write better code.

## Philosophy behind the tools of R Package development

* Anything that can be automated, should be automated. 
* Use functions for package development, avoid doing it by hand. 
   - To do this, we need to learn the functions/workflow. 
   - Our thinking should be on what we want the package to do, not on how to organize a package. 
   - This is where the [`devtools`](https://devtools.r-lib.org/) and [`usethis`](https://usethis.r-lib.org/) package come into play. 
   - Tools to insulate us from the low-level details of how packages are built.
* `devtools` works nicely with RStudio, so that's the development environment we will use. 

```{r 01-install-devtools-usethis, eval=FALSE}
install.packages(c("devtools", "usethis"))
```

## What if we need more detail?

* The official manual is [Writing R Extensions](https://cran.r-project.org/doc/manuals/R-exts.html).
   - This document is exhaustive. 

## A note 

* Don't get overwhelmed
  - Some stuff you may never use.
  - Remember, an R package at it's core is just an `R/` directory.
  
* You are not expected to know everything.
  - We are all learning.
  - None of us is as smart as all of us.
  - It's okay to say "I don't know."
  - We can stop at anytime to discuss.


## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/QOmdmGMxl_Q")`

<details>
<summary> Meeting chat log </summary>

```
00:07:46	Rex:	Brisbane, Australia
00:36:26	Collin Berke:	On a scale from 1 - 5, how familiar are you with using git for version control?
00:36:51	Ryan Metcalf:	4
00:36:52	Brendan Lam:	2
00:36:52	Arun Chavan:	3
00:36:53	Larissa Shaughnessy:	2
00:36:54	Isabella Velásquez:	4
00:36:57	Rex Parsons:	3
00:37:19	Collin Berke:	On a scale from 1 - 5, how familiar are you with using GitHub?
00:37:38	Rex Parsons:	4
00:37:38	Arun Chavan:	4
00:37:38	Isabella Velásquez:	5
00:37:39	Brendan Lam:	2
00:37:41	Larissa Shaughnessy:	3
00:38:14	Collin Berke:	On a scale from 1 - 5, how valuable would it be to devote a week to introducing git and GitHub?
00:38:29	Brendan Lam:	3
00:38:31	Arun Chavan:	2
00:38:35	Larissa Shaughnessy:	4
00:38:35	Ryan Metcalf:	4
00:38:36	Rex Parsons:	4
00:38:48	Isabella Velásquez:	I am comfortable with Git, but if it’s valuable for the group!
01:05:36	Rex Parsons:	Thanks, Collin!
```
</details>
