# System setup

**Learning objectives:**

- How to check which version of R and RStudio you are using
- Discover `devtools`
- Learn about the `{usethis}` package
- Understand the distict difference between useRs and developeRs

## Prepare your system

To start the preparation of development environment, lets get our bearings.

```{r, How to check your R version, eval=TRUE}
R.Version()

```

```{r, Another way to check your R version, eval=TRUE}
sessionInfo()
```

Ok, that was fun, but the version check isn't in the book! We need to follow the author, right?

Let's load some packages to prepare authoring some packages!

```{r, Prepare your system, eval=FALSE}
install.packages(c("devtools", "roxygen2", "testthat", "knitr"))
```

> NOTE: For some reason, this script caused an infinite loop of restart R, attempt to install, request to restart R, attempt to install...infinity. If this should happen, select *No*.

The author recommends increasing to the *Preview* version of RStudio Integrated Development Environment (IDE). We are considering outselves developers, right?

- Usually this isn't needed or required for common users (we're not common)
- It puts you in a position to be subject to bugs (not as many as the daily builds though)
- [Preview version](https://www.rstudio.com/products/rstudio/download/preview/)
- [Released version]( https://www.rstudio.com/products/rstudio/download/)

## devtools, usethis, and you

At some point in your development career...you will find you need to reorder to progress further. This is exactly what happened to the `{devtools}` package. In 2018, it was reordered and split into seven sub-packages:

- `{remotes}`
- `{pkgbuild}`
- `{pkgload}`
- `{rcmdcheck}`
- `{revdepcheck}`
- `{sessioninfo}`
- `{usethis}`

What is cool here, we already used one of these packages!!!

You can think of devtools as a wrapper providing the ability to:

- set user-friendly defaults
- introduce helpful interactive behaviour
- combine functionality from multiple sub-packages

There is a difference in useRs and developeRs. They are:

- useRs should attach devtools and think of it as the provider of your favorite functions for package development
- developeRs should **NOT** depend on devtools, but should instead access functions via the package that is their primary home
  - devtools should rarely appear in the role of `foo` in a qualified call of the form `foo::fcn()`. Instead, `foo` should be the package where `fcn()` is defined.
  - An exception to this is that we continue to feature `devtools::install_github()` as the way to install the development version of a package in its README
- If required, ensure you report bugs to each sub-package

Example of how to simulate installing and loading a package (interactive development)

```{r, Interactive pacakge load, eval=TRUE}
library(devtools)
load_all()
```

If used inside an R package, this is the preferred call:

```{r, Inside an R Package, eval=TRUE}
pkgload::load_all()
```


### Personal startup configuration
Attach devtools with the following script:

```{r, Installing DevTools, eval=TRUE}
library(devtools)
```

Doing this everytime you restart your session gets old...quick! Lets be more efficient with our keystrokes.

Let's edit our `.Rprofile` file. If it does not exist on your computer, you can run `use_devtools()` to create it.

The output of my command looks like this:

```{r, Example .Rprofile entry, eval=FALSE}
use_devtools()
• Include this code in '.Rprofile' to make devtools available in all interactive sessions.
  if (interactive()) {
    suppressMessages(require(devtools))
  }
  [Copied to clipboard]
• Modify '/Users/rmetcalf/.Rprofile'
• Restart R for changes to take effect
```

Caution, Warning, HUGE RED FLAG!!!! You are not a Super-person yet!

In general, it’s a bad idea to attach packages in .Rprofile, as it invites you to create R scripts that don’t reflect all of their dependencies via explicit calls to library(foo). But devtools is a workflow package that smooths the process of package development and is, therefore, unlikely to get baked into any analysis scripts.

> Note how we still take care to only attach in interactive sessions.

I may be too harsh. The following is a good example when `usethis` helps and can be added to your .Rprofile.

```{r, Example R Code, eval=FALSE}
options(
  usethis.full_name = "Jane Doe",
  usethis.description = list(
    `Authors@R` = 'person("Jane", "Doe", email = "jane@example.com", role = c("aut", "cre")), 
    comment = c(ORCID = "YOUR-ORCID-ID"))',
    License = "MIT + file LICENSE",
    Version = "0.0.0.9000"
  ),
  usethis.protocol  = "ssh"  
)
```

For history purposes (or at some point it may change....)

```{r, Install DevTools and UseThis, eval=TRUE}
devtools::install_github("r-lib/devtools")
devtools::install_github("r-lib/usethis")
```

## R build toolchain

Any good developer must have a good compiler. The following three sections will expand on how this may be accomplished via the three main operating systems.

>NOTE: At this point, these may not be required. The compiler options will becomre important when a source code contains C or C++ code.

### Windows

On Windows the collection of tools needed for building packages from source is called [Rtools](https://cran.r-project.org/bin/windows/Rtools/).

Rtools is **NOT** an R package and therefore cannot be installed using `install.packages('Rtools')`

During the Rtools installation you may see a window asking you to *Select Additional Tasks*.

- **DO NOT** select the box for ***Edit the system PATH***.
- **DO** select the box for ***Save version information to registry***.

### macOS

If development is on a Mac, you will require Xcode.

In shell/terminal run the following code:

```{bash, eval=FALSE}
xcode-select --install
```

>NOTE: Xcode is large....like, ~90GBs large (an estimate with all simulators)! There are options to reduce the size and can be found [here](https://apple.stackexchange.com/questions/287307/reduce-size-of-the-xcode-application#:~:text=Go%20to%20~%2FLibrary%2FDeveloper,%2FXcode%2FiOS%20DeviceSupport%2F%20.).

Alternatively, you can also download Xcode from the [Mac App Store](https://itunes.apple.com/ca/app/xcode/id497799835?mt=12).

### Linux

Linux install is the simplest (dependent on your distro of choice and package manager).

Ensure you include the following:

- R
- R development tools

For example, on Ubuntu (and Debian):

```{bash, eval=FALSE}
sudo apt-get update & upgrade
sudo apt-get install r-base-dev package
```

Other Linux distributions may have alternative commands.

### Verify system prep

To validate everying is installed or up to date, run the following:

```{r, Validation and Update if required, eval=TRUE}
devtools::dev_sitrep()
```

Update any packages based on recommendations.
