
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```


# The package within

**Purpose**

To walk through the development of a small toy package, with an emphasis on the __package's__ R code and how it differs from R code in a __script__.

Roadmap:

1. Data analysis script 
2. Isolate and extract the reusable data and logic from the script 
3. Put code into an R package 
4. Use package in a newly simplified script.

This chapter purposefully makes mistakes along the way, in order to highlight special considerations for the R code inside a package.

*Section headers incorporate the NATO phonetic alphabet.*

## Alfa: a script that works

A fictional script for a dataset of people who went for a swim:

> Where did you swim and how hot was it outside?

```{r}
infile <- "swim.csv"
(dat <- read.csv(infile))
#>   name    where temp
#> 1 Adam    beach   95
#> 2 Bess    coast   91
#> 3 Cora seashore   28
#> 4 Dale    beach   85
#> 5 Evan  seaside   31
```

1. Observations were classified as American or British based on how they described the beach:

```{r}
dat$english[dat$where == "beach"] <- "US"
dat$english[dat$where == "coast"] <- "US"
dat$english[dat$where == "seashore"] <- "UK"
dat$english[dat$where == "seaside"] <- "UK"
```

2. Temperatures were converted to Celsius:

```{r}
dat$temp[dat$english == "US"] <- (dat$temp[dat$english == "US"] - 32) * 5/9
dat
#>   name    where temp english
#> 1 Adam    beach 35.0      US
#> 2 Bess    coast 32.8      US
#> 3 Cora seashore 28.0      UK
#> 4 Dale    beach 29.4      US
#> 5 Evan  seaside 31.0      UK
```

3. Data is written back out to a CSV file. A timestamp is also captured in the filename.

```{r}
now <- Sys.time()
timestamp <- format(now, "%Y-%B-%d_%H-%M-%S")
(outfile <- paste0(timestamp, "_", sub("(.*)([.]csv$)", "\\1_clean\\2", infile)))
#> [1] "2022-April-17_07-14-16_swim_clean.csv"
write.csv(dat, file = outfile, quote = FALSE, row.names = FALSE)
```

## Bravo: a better script that works

- There's a package that lurks within the original script.
    - Suboptimal coding practices like repetitive code and mixing of code and data makes it hard to catch.
- A good first step is to refactor this code.
    
Next version of the script:

```{r}
library(tidyverse)

infile <- "swim.csv"
dat <- read_csv(infile, col_types = cols(name = "c", where = "c", temp = "d"))

lookup_table <- tribble(
      ~where, ~english,
     "beach",     "US",
     "coast",     "US",
  "seashore",     "UK",
   "seaside",     "UK"
)

dat <- dat %>% 
  left_join(lookup_table)
#> Joining, by = "where"

f_to_c <- function(x) (x - 32) * 5/9

dat <- dat %>% 
  mutate(temp = if_else(english == "US", f_to_c(temp), temp))
dat
#> # A tibble: 5 × 4
#>   name  where     temp english
#>   <chr> <chr>    <dbl> <chr>  
#> 1 Adam  beach     35   US     
#> 2 Bess  coast     32.8 US     
#> 3 Cora  seashore  28   UK     
#> 4 Dale  beach     29.4 US     
#> 5 Evan  seaside   31   UK

now <- Sys.time()
timestamp <- function(time) format(time, "%Y-%B-%d_%H-%M-%S")
outfile_path <- function(infile) {
  paste0(timestamp(now), "_", sub("(.*)([.]csv$)", "\\1_clean\\2", infile))
}
write_csv(dat, outfile_path(infile))
```

**Key features of this code:**

- Functions from the `tidyverse` packages, `readr` and `dplyr`.
- Different “beach” words are stored in a _lookup table_. This makes it easier to add words in the future.
- `f_to_c()`, `timestamp()`, and `outfile_path()` functions now hold the logic for converting temperatures and forming the timestamped output file name.


Easier to recognize the reusable bits of this script, i.e. the bits that have nothing to do with a specific input file, like `swim.csv`

## Charlie: external helpers

Move reusable data and logic out of the analysis script and into separate files.

Here is the content of `beach-lookup-table.csv`:

```
where,english
beach,US
coast,US
seashore,UK
seaside,UK
```

Here is the content of `cleaning-helpers.R`:

```{r}
library(tidyverse)

localize_beach <- function(dat) {
  lookup_table <- read_csv(
    "beach-lookup-table.csv",
    col_types = cols(where = "c", english = "c")
  )
  left_join(dat, lookup_table)
}

f_to_c <- function(x) (x - 32) * 5/9

celsify_temp <- function(dat) {
  mutate(dat, temp = if_else(english == "US", f_to_c(temp), temp))
}

now <- Sys.time()
timestamp <- function(time) format(time, "%Y-%B-%d_%H-%M-%S")
outfile_path <- function(infile) {
  paste0(timestamp(now), "_", sub("(.*)([.]csv$)", "\\1_clean\\2", infile))
}
```

- High-level helper functions like `localize_beach()` and `celsify_temp()`, were added to the pre-existing helpers (`f_to_c()`, `timestamp()`, and `outfile_path()`)

```{r}
library(tidyverse)
source("cleaning-helpers.R")

infile <- "swim.csv"
dat <- read_csv(infile, col_types = cols(name = "c", where = "c", temp = "d"))

(dat <- dat %>% 
    localize_beach() %>% 
    celsify_temp())
#> Joining, by = "where"
#> # A tibble: 5 × 4
#>   name  where     temp english
#>   <chr> <chr>    <dbl> <chr>  
#> 1 Adam  beach     35   US     
#> 2 Bess  coast     32.8 US     
#> 3 Cora  seashore  28   UK     
#> 4 Dale  beach     29.4 US     
#> 5 Evan  seaside   31   UK

write_csv(dat, outfile_path(infile))
```

Script is now much shorter (and cleaner). However, whether it's easier depends on personal preference and what "feels" easier to work with.

For more information, refer to the [Tidyverse design guide](https://design.tidyverse.org/)


## Delta: an attempt at a package

- Use `usethis::create_package()` to scaffold a new R package
- Copy `cleaning-helpers.R` into the new package, specifically, to `R/cleaning-helpers.R`
- Copy `beach-lookup-table.csv` into the top-level of the new source package
- Install package

Script we're trying to run:

```{r}
library(tidyverse)
library(delta) # instead of source("cleaning-helpers.R")

infile <- "swim.csv"
dat <- read_csv(infile, col_types = cols(name = "c", where = "c", temp = "d"))

dat <- dat %>% 
  localize_beach() %>% 
  celsify_temp()

write_csv(dat, outfile_path(infile))
```

Results when we try to run this code:

```{r}
library(tidyverse)
library(delta)

infile <- "swim.csv"
dat <- read_csv(infile, col_types = cols(name = "c", where = "c", temp = "d"))

dat <- dat %>% 
  localize_beach() %>% 
  celsify_temp()
> Error in localize_beach(.) : could not find function "localize_beach"

write_csv(dat, outfile_path(infile))
> Error in outfile_path(infile) : could not find function "outfile_path"
```

Despite calling `library(delta)`, none of the functions were actually available to use.

- Attaching a package does not put the functions in the global workspace.
      - In contrast to `source()`-ing a file of helper functions, attaching a package does not dump its functions into the global workspace. 
      - By default, functions in a package are only for internal use
      - We can export these functions properly by putting `@export` in the roxygen comment above each function (See Section 10.4 for more information on DESCRIPTION)

```{r}
#' @export
celsify_temp <- function(dat) {
  mutate(dat, temp = if_else(english == "US", f_to_c(temp), temp))
}
```

Now our script works (sort of)!

```{r}
library(tidyverse)
library(delta)

infile <- "swim.csv"
dat <- read_csv(infile, col_types = cols(name = "c", where = "c", temp = "d"))

dat <- dat %>% 
  localize_beach() %>% 
  celsify_temp()
#> Error: 'beach-lookup-table.csv' does not exist in current working directory ('/Users/jenny/tmp').

write_csv(dat, outfile_path(infile))
```

Problem: You can't dump CSV files into the source of an R package and expect it to work. Despite this, you can still install and attach this package.

- This means that broken packages can still be used. To prevent this, you should run `R CMD` check or `check()` often during development.
   - Doing so will alert you to the problem:

```
* installing *source* package ‘delta’ ...
** using staged installation
** R
** byte-compile and prepare package for lazy loading
Error in library(tidyverse) : there is no package called ‘tidyverse’
Error: unable to load R code in package ‘delta’
Execution halted
ERROR: lazy loading failed for package ‘delta’
* removing ‘/Users/brendan/RScripts/delta.Rcheck/delta’
```

What is the reason behind this error?

- Package was declared incorrectly
   - While you can load a package using `library(tidyverse)` in a script, dependencies on other packages must be declared in the `DESCRIPTION` 
   
   
## Echo: a working package

Now we'll make a package that actually works:

```{r}
lookup_table <- dplyr::tribble(
      ~where, ~english,
     "beach",     "US",
     "coast",     "US",
  "seashore",     "UK",
   "seaside",     "UK"
)

#' @export
localize_beach <- function(dat) {
  dplyr::left_join(dat, lookup_table)
}

f_to_c <- function(x) (x - 32) * 5/9

#' @export
celsify_temp <- function(dat) {
  dplyr::mutate(dat, temp = dplyr::if_else(english == "US", f_to_c(temp), temp))
}

now <- Sys.time()
timestamp <- function(time) format(time, "%Y-%B-%d_%H-%M-%S")

#' @export
outfile_path <- function(infile) {
  paste0(timestamp(now), "_", sub("(.*)([.]csv$)", "\\1_clean\\2", infile))
}
```

**Note:** To fix our initial problem with loading a CSV file, we've created a dataframe `lookup_table`. However, Chapter 8 Data provides more guidance and recommendations on how to load datasets properly.

**Other Note:** When calling functions from other packages, we should specify the package that we're using (e.g., dplyr::mutate()). Moreover, we should identify the specific package being used, rather than the meta-package (e.g., do not use tidyverse::mutate())

- All of the user-facing functions have an `@export` tag in their roxygen comment, which means that devtools::document() adds them correctly to the NAMESPACE file.

This package can be installed, but we receive 1 note and 1 warning:

``` 
* checking R code for possible problems ... NOTE
celsify_temp: no visible binding for global variable ‘english’
celsify_temp: no visible binding for global variable ‘temp’
Undefined global functions or variables:
  english temp

* checking for missing documentation entries ... WARNING
Undocumented code objects:
  ‘celsify_temp’ ‘localize_beach’ ‘outfile_path’
All user-level objects in a package should have documentation entries.
See chapter ‘Writing R documentation files’ in the ‘Writing R
Extensions’ manual.
```
Translation of first warnings: `no visible binding for global variable ‘english’` and `no visible binding for global variable ‘temp’`

- Using bare variable names like `english` and `temp` looks suspicious because you're using unquoted variable names from `dplyr` inside a package.
   - Defining these variables globally eliminates the note:

```{r}
option 1 (then you should also put utils in Imports)
utils::globalVariables(c("english", "temp"))

option 2
english <- temp <- NULL
```
The other note we received from R: 

`"Undocumented code objects: ‘celsify_temp’ ‘localize_beach’ ‘outfile_path’ All user-level objects in a package should have documentation entries."` 

This is caused by not documenting exported functions. Using roxygen comments to document it should solve the problem. (Function documentation is discussed in Chapter 16).


## Foxtrot: build time vs. run time

**Another problem::** timestamps don't seem to work properly.

```{r}
Sys.time()
#> [1] "2022-02-24 20:49:59 PST"

outfile_path("INFILE.csv")
#> [1] "2020-September-03_11-06-33_INFILE_clean.csv"
```
The timestamp reflects the time that the function was initially run, rather than the current time.

Recall how we form the filepath for output files:

```{r}
now <- Sys.time()
timestamp <- function(time) format(time, "%Y-%B-%d_%H-%M-%S")
outfile_path <- function(infile) {
  paste0(timestamp(now), "_", sub("(.*)([.]csv$)", "\\1_clean\\2", infile))
}
```


**Source of the problem:** The `now <- Sys.time()` function lies outside the `outfile_path` definition, so it is executed when the package is built, but never again. *Code outside your functions is only built once at build time.*

Moving `now <- Sys.time` so that it's no longer top level code:

```{r}
# always timestamp as "now"
outfile_path <- function(infile) {
  now <- timestamp(Sys.time())
  paste0(now, "_", sub("(.*)([.]csv$)", "\\1_clean\\2", infile))
}

# allow user to provide a time, but default to "now"
outfile_path <- function(infile, time = Sys.time()) {
  now <- timestamp(time)
  paste0(now, "_", sub("(.*)([.]csv$)", "\\1_clean\\2", infile))
}
```

> Need to have a different mindset when defining objects inside a package. The objects should be functions and these functions should (generally) only use data they create or that is passed via an argument.


## Golf: side effects

A new concern with the timestamp: The timestamps depend on which part of the world you're in.
```{r, results='hide', warning=FALSE, echo=FALSE}
suppressPackageStartupMessages(library(tidyverse))
```


```{r echo=FALSE}
location <- c("Rome, Italy", "Warsaw, Poland", "Sao Paulo, Brazil", "Greenwich, England")
timestamp <- c("2020-settembre-05_00-30-00", "2020-września-05_00-30-00", "2020-setembro-04_19-30-00", "2020-September-04_23-30-00")
LC_TIME <- c("it_IT.UTF-8", "pl_PL.UTF-8", "pt_BR.UTF-8", "en_GB.UTF-8")
tz <- c("Europe/Rome", "Europe/Warsaw", "America/Sao_Paulo", "Europe/London")

tibble::tibble(location, timestamp, LC_TIME, tz) %>% 
  knitr::kable()
```

**Problem:** The month names vary, as does the time, and even the date.

**Proposed Solution:** Create timestamps that are all in a fixed time zone.

- We can force a certain locale with `Sys.setlocale()` and force a particular time zone by adjusting the TZ environment variable.

Our attempt at implementing this:
```
timestamp <- function(time = Sys.time()) {
  Sys.setlocale("LC_TIME", "C")
  Sys.setenv(TZ = "UTC")
  format(time, "%Y-%B-%d_%H-%M-%S")
}
```
However, a user in Brazil would see this after using `outfile_path()` from our package:
```
outfile_path("INFILE.csv")
#> [1] "2022-April-17_07-14-18_INFILE_clean.csv"

format(Sys.time(), "%Y-%B-%d_%H-%M-%S")
#> [1] "2022-April-17_07-14-18"
```

Our calls to `Sys.setlocale()` and `Sys.setenv()` inside `timestamp()` have made persistent changes to their R session. This sort of side effect is very undesirable and is extremely difficult to track down and debug, especially in more complicated settings.

**Solution:**
```
# use withr::local_*() functions to keep the changes local to timestamp()
timestamp <- function(time = Sys.time()) {
  withr::local_locale(c("LC_TIME" = "C"))
  withr::local_timezone("UTC")
  format(time, "%Y-%B-%d_%H-%M-%S")
}

# use the tz argument to format.POSIXct()
timestamp <- function(time = Sys.time()) {
  withr::local_locale(c("LC_TIME" = "C"))
  format(time, "%Y-%B-%d_%H-%M-%S", tz = "UTC")
}

# put the format() call inside withr::with_*()
timestamp <- function(time = Sys.time()) {
  withr::with_locale(
    c("LC_TIME" = "C"),
    format(time, "%Y-%B-%d_%H-%M-%S", tz = "UTC")
  )
}
```
The locale in our timestamp is only temporarily modified with the `withr::with_locale` function.

- In this example, the mistake we made was changing the user's overall state. 
   - If you have to do this, make sure this is documented explicitly or try to make them reversible


> You need to adopt a different mindset when defining functions inside a package. Try to avoid making any changes to the user’s overall state. If such changes are unavoidable, make sure to reverse them (if possible) or to document them explicitly (if related to the function’s primary purpose).

## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/eMWgu9OQ0m8")`


### Cohort 2

`r knitr::include_url("https://www.youtube.com/embed/yRzRiXiqPag")`


### Cohort 3

`r knitr::include_url("https://www.youtube.com/embed/bsGCgJr60as")`

<details>
<summary> Meeting chat log </summary>

```
00:45:32	collinberke:	Check this out regarding side effects: https://withr.r-lib.org/articles/changing-and-restoring-state.html
00:55:58	Arun Chavan:	https://rstats.wtf/
00:58:05	Isabella Velásquez:	https://yihui.org/knitr/options/
00:59:31	Arun Chavan:	+1 for knitr options because the tab autocompletion doesn’t work for them (at least for me)
01:00:58	Brendan Lam:	I'm gonna try that
```
</details>


### Cohort 4

`r knitr::include_url("https://www.youtube.com/embed/xHYtYBrNAbg")`

<details>
<summary> Meeting chat log </summary>

```
00:25:03	Olivier Leroy:	dplyr:::select()
00:26:06	Olivier Leroy:	dplyr::select()
00:28:56	Olivier Leroy:	https://docs.google.com/spreadsheets/d/1AlgFwMR42NjjoG2lcm0hf_pk5ZACw-n8UCwnrny91bU/edit#gid=0
00:35:45	Howard Baek:	https://github.com/r-lib/conflicted
```
</details>
