# The package within

**Purpose**

To walk through the building of a toy package, with an emphasis on the __package's__ R code and how it differs from R code in a __script__.

Data analysis script $\rightarrow$ Isolate and extract the reusable data and logic from the script $\rightarrow$ Put code into an R package $\rightarrow$ Use package in a newly simplified script    


## Alfa: a script that works

A fictional script for a dataset of people who went for a swim.

> Where did you swim and how hot was it outside?

```
infile <- "swim.csv"
(dat <- read.csv(infile))
#>   name    where temp
#> 1 Adam    beach   95
#> 2 Bess    coast   91
#> 3 Cora seashore   28
#> 4 Dale    beach   85
#> 5 Evan  seaside   31
```

1. Observations were classified as American or British based on how they described the beach:
```
dat$english[dat$where == "beach"] <- "US"
dat$english[dat$where == "coast"] <- "US"
dat$english[dat$where == "seashore"] <- "UK"
dat$english[dat$where == "seaside"] <- "UK"
```
2. Temperatures were converted to Celsius:
```
dat$temp[dat$english == "US"] <- (dat$temp[dat$english == "US"] - 32) * 5/9
dat
#>   name    where temp english
#> 1 Adam    beach 35.0      US
#> 2 Bess    coast 32.8      US
#> 3 Cora seashore 28.0      UK
#> 4 Dale    beach 29.4      US
#> 5 Evan  seaside 31.0      UK
```
3. Data is written back to a csv file. A timestamp is also captured in the filename.
```
now <- Sys.time()
timestamp <- format(now, "%Y-%B-%d_%H-%M-%S")
(outfile <- paste0(timestamp, "_", sub("(.*)([.]csv$)", "\\1_clean\\2", infile)))
#> [1] "2022-April-17_07-14-16_swim_clean.csv"
write.csv(dat, file = outfile, quote = FALSE, row.names = FALSE)
```

## Bravo: a better script that works
- There's a package that lurks within the original script (apparently)
    - Suboptimal coding practices like repetitive code and mixing of code and data
    
Next version of the script:
```
library(tidyverse)

infile <- "swim.csv"
dat <- read_csv(infile, col_types = cols(name = "c", where = "c", temp = "d"))

lookup_table <- tribble(
      ~where, ~english,
     "beach",     "US",
     "coast",     "US",
  "seashore",     "UK",
   "seaside",     "UK"
)

dat <- dat %>% 
  left_join(lookup_table)
#> Joining, by = "where"

f_to_c <- function(x) (x - 32) * 5/9

dat <- dat %>% 
  mutate(temp = if_else(english == "US", f_to_c(temp), temp))
dat
#> # A tibble: 5 × 4
#>   name  where     temp english
#>   <chr> <chr>    <dbl> <chr>  
#> 1 Adam  beach     35   US     
#> 2 Bess  coast     32.8 US     
#> 3 Cora  seashore  28   UK     
#> 4 Dale  beach     29.4 US     
#> 5 Evan  seaside   31   UK

now <- Sys.time()
timestamp <- function(time) format(time, "%Y-%B-%d_%H-%M-%S")
outfile_path <- function(infile) {
  paste0(timestamp(now), "_", sub("(.*)([.]csv$)", "\\1_clean\\2", infile))
}
write_csv(dat, outfile_path(infile))
```

**Key features of this code:**

- using functions from tidyverse packages
- different “beach” words are stored in a lookup table. This makes it easier to add words in the future.
- `f_to_c()`, `timestamp()`, and `outfile_path()` functions now hold the logic for converting temperatures and forming the timestamped output file name.

## Charlie: external helpers
Move reusable data and logic out of the analysis script and into separate files.

```
library(tidyverse)

localize_beach <- function(dat) {
  lookup_table <- read_csv(
    "beach-lookup-table.csv",
    col_types = cols(where = "c", english = "c")
  )
  left_join(dat, lookup_table)
}

f_to_c <- function(x) (x - 32) * 5/9

celsify_temp <- function(dat) {
  mutate(dat, temp = if_else(english == "US", f_to_c(temp), temp))
}

now <- Sys.time()
timestamp <- function(time) format(time, "%Y-%B-%d_%H-%M-%S")
outfile_path <- function(infile) {
  paste0(timestamp(now), "_", sub("(.*)([.]csv$)", "\\1_clean\\2", infile))
}
```

- High-level helper functions like localize_beach() and celsify_temp(), were added to the pre-existing helpers (f_to_c(), timestamp(), and outfile_path())

```
library(tidyverse)
source("cleaning-helpers.R")

infile <- "swim.csv"
dat <- read_csv(infile, col_types = cols(name = "c", where = "c", temp = "d"))

(dat <- dat %>% 
    localize_beach() %>% 
    celsify_temp())
#> Joining, by = "where"
#> # A tibble: 5 × 4
#>   name  where     temp english
#>   <chr> <chr>    <dbl> <chr>  
#> 1 Adam  beach     35   US     
#> 2 Bess  coast     32.8 US     
#> 3 Cora  seashore  28   UK     
#> 4 Dale  beach     29.4 US     
#> 5 Evan  seaside   31   UK

write_csv(dat, outfile_path(infile))
```
Script is now much shorter (and cleaner). However, whether it's easier depends on personal preference and what "feels" easier to work with.

## Delta: an attempt at a package

- Use usethis::create_package() to scaffold a new R package
- Copy cleaning-helpers.R into the new package, specifically, to R/cleaning-helpers.R
- Copy beach-lookup-table.csv into the top-level of the new source package
- install package

Script we're trying to run:
```
library(tidyverse)
library(delta) # originally source("cleaning-helpers.R")

infile <- "swim.csv"
dat <- read_csv(infile, col_types = cols(name = "c", where = "c", temp = "d"))

dat <- dat %>% 
  localize_beach() %>% 
  celsify_temp()

write_csv(dat, outfile_path(infile))
```
Results when we try to run this code:
```
library(tidyverse)
library(delta)

infile <- "swim.csv"
dat <- read_csv(infile, col_types = cols(name = "c", where = "c", temp = "d"))

dat <- dat %>% 
  localize_beach() %>% 
  celsify_temp()
> Error in localize_beach(.) : could not find function "localize_beach"

write_csv(dat, outfile_path(infile))
> Error in outfile_path(infile) : could not find function "outfile_path"
```

Despite calling our library(delta), none of the functions were actually available to use.

- This is because we only attached the package, rather than sourcing the file with helper functions. 
   - Attaching a package does not put the functions in the global workspace.
      - We can export these functions properly by putting `@export` in the roxygen comment above each function

```
#' @export
celsify_temp <- function(dat) {
  mutate(dat, temp = if_else(english == "US", f_to_c(temp), temp))
}
```

Now our script works (sort of)!

```
library(tidyverse)
library(delta)

infile <- "swim.csv"
dat <- read_csv(infile, col_types = cols(name = "c", where = "c", temp = "d"))

dat <- dat %>% 
  localize_beach() %>% 
  celsify_temp()
#> Error: 'beach-lookup-table.csv' does not exist in current working directory ('/Users/jenny/tmp').

write_csv(dat, outfile_path(infile))
```

Problem: You can't dump CSV files into the source of an R package and expect it to work. Despite this, you can still install and attach this package.

- This means that broken packages can still be used. To prevent this, you should run `R CMD` check or `check()` often during development.
   - Doing so will alert you to the problem:

```
* installing *source* package ‘delta’ ...
** using staged installation
** R
** byte-compile and prepare package for lazy loading
Error in library(tidyverse) : there is no package called ‘tidyverse’
Error: unable to load R code in package ‘delta’
Execution halted
ERROR: lazy loading failed for package ‘delta’
* removing ‘/Users/brendan/RScripts/delta.Rcheck/delta’
```

What are the reasons behing these errors?

- Package was declared incorrectly
   - While you can load a package using `library(tidyverse)` in a script, dependencies on other packages must be declared in the `DESCRIPTION` 
   
## Echo: a working package
Now we'll make a package that actually works:
```
lookup_table <- dplyr::tribble(
      ~where, ~english,
     "beach",     "US",
     "coast",     "US",
  "seashore",     "UK",
   "seaside",     "UK"
)

#' @export
localize_beach <- function(dat) {
  dplyr::left_join(dat, lookup_table)
}

f_to_c <- function(x) (x - 32) * 5/9

#' @export
celsify_temp <- function(dat) {
  dplyr::mutate(dat, temp = dplyr::if_else(english == "US", f_to_c(temp), temp))
}

now <- Sys.time()
timestamp <- function(time) format(time, "%Y-%B-%d_%H-%M-%S")

#' @export
outfile_path <- function(infile) {
  paste0(timestamp(now), "_", sub("(.*)([.]csv$)", "\\1_clean\\2", infile))
}
```

**Note:** To fix our initial problem with loading a CSV file, we've created lookup_table to create a data frame instead. However, Chapter 14 provides more guidance and recommendations on how to load datasets properly.

**Other Note:** When calling functions from other packages, we should specify the package that we're using (e.g., dplyr::mutate()). Moreover, we should identify the specific package being used, rather than the meta-package (e.g., do not use tidyverse::mutate())

- All of the user-facing functions have an `@export` tag in their roxygen comment, which means that devtools::document() adds them correctly to the NAMESPACE file.

This package can be installed, but we receive 1 note and 1 warning:
``` 
* checking R code for possible problems ... NOTE
celsify_temp: no visible binding for global variable ‘english’
celsify_temp: no visible binding for global variable ‘temp’
Undefined global functions or variables:
  english temp

* checking for missing documentation entries ... WARNING
Undocumented code objects:
  ‘celsify_temp’ ‘localize_beach’ ‘outfile_path’
All user-level objects in a package should have documentation entries.
See chapter ‘Writing R documentation files’ in the ‘Writing R
Extensions’ manual.
```
Translation of first warnings: `no visible binding for global variable ‘english’` and `no visible binding for global variable ‘temp’`

- Using bare variable names like english and temp looks suspicious because you're using unquoted variable names from dplyr inside a package.
   - defining these variables globally eliminates the note:

```
option 1 (then you should also put utils in Imports)
utils::globalVariables(c("english", "temp"))

option 2
english <- temp <- NULL
```
The other note we received from R: 

`"Undocumented code objects: ‘celsify_temp’ ‘localize_beach’ ‘outfile_path’ All user-level objects in a package should have documentation entries."` 

This is caused by not documenting exported functions. Using roxygen comments to document it should solve the problem.

## Foxtrot: build time vs. run time

Another problem: For this package, the timestamps don't seem to work properly..
```
Sys.time()
#> [1] "2022-02-24 20:49:59 PST"

outfile_path("INFILE.csv")
#> [1] "2020-September-03_11-06-33_INFILE_clean.csv"
```
The timestamp reflects the time that the function was initially run, rather than the current time.

**Source of the problem:** The `Sys.time()` function outside the `outfile_path` definition. The `Sys.time` is executed when the package is built, but never again. *Code outside your functions is only built once at build time.*

Moving `Sys.time` so that it's no longer top level code:
```
# always timestamp as "now"
outfile_path <- function(infile) {
  ts <- timestamp(Sys.time())
  paste0(ts, "_", sub("(.*)([.]csv$)", "\\1_clean\\2", infile))
}

# allow user to provide a time, but default to "now"
outfile_path <- function(infile, time = Sys.time()) {
  ts <- timestamp(time)
  paste0(ts, "_", sub("(.*)([.]csv$)", "\\1_clean\\2", infile))
}
```

**Takeaway:** Need to have a different mindset when defining objects for a package. The objects should be functions and these functions should (generally) only use data they create or that is passed via an argument.

## Golf: side effects

A new concern with the timestamp: The timestamps depend on which part of the world you're in.
```{r, results='hide', warning=FALSE, echo=FALSE}
suppressPackageStartupMessages(library(tidyverse))
```


```{r echo=FALSE}
location <- c("Rome, Italy", "Warsaw, Poland", "Sao Paulo, Brazil", "Greenwich, England")
timestamp <- c("2020-settembre-05_00-30-00", "2020-września-05_00-30-00", "2020-setembro-04_19-30-00", "2020-September-04_23-30-00")
LC_TIME <- c("it_IT.UTF-8", "pl_PL.UTF-8", "pt_BR.UTF-8", "en_GB.UTF-8")
tz <- c("Europe/Rome", "Europe/Warsaw", "America/Sao_Paulo", "Europe/London")

tibble::tibble(location, timestamp, LC_TIME, tz) %>% 
  knitr::kable()
```

**Proposed Solution:** Create timestamps that are all in a fixed time zone.

- We can force a certain locale with `Sys.setlocale()` and force a particular time zone by adjusting the TZ environment variable.

Our attempt at implementing this:
```
timestamp <- function(time = Sys.time()) {
  Sys.setlocale("LC_TIME", "C")
  Sys.setenv(TZ = "UTC")
  format(time, "%Y-%B-%d_%H-%M-%S")
}
```
However, a user in Brazil would see this after using `outfile_path()` from our package:
```
outfile_path("INFILE.csv")
#> [1] "2022-April-17_07-14-18_INFILE_clean.csv"

format(Sys.time(), "%Y-%B-%d_%H-%M-%S")
#> [1] "2022-April-17_07-14-18"
```

Our calls to `Sys.setlocale()` and `Sys.setenv()` inside `timestamp()` have made persistent changes to their R session. This sort of side effect is very undesirable and is extremely difficult to track down and debug, especially in more complicated settings.

**Solution:**
```
# use withr::local_*() functions to keep the changes local to timestamp()
timestamp <- function(time = Sys.time()) {
  withr::local_locale(c("LC_TIME" = "C"))
  withr::local_timezone("UTC")
  format(time, "%Y-%B-%d_%H-%M-%S")
}

# use the tz argument to format.POSIXct()
timestamp <- function(time = Sys.time()) {
  withr::local_locale(c("LC_TIME" = "C"))
  format(time, "%Y-%B-%d_%H-%M-%S", tz = "UTC")
}

# put the format() call inside withr::with_*()
timestamp <- function(time = Sys.time()) {
  withr::with_locale(
    c("LC_TIME" = "C"),
    format(time, "%Y-%B-%d_%H-%M-%S", tz = "UTC")
  )
}
```
The locale in our timestamp is only temporarily modified with the `withr::with_locale` function.

- In this example, the mistake we made was changing the user's overall state. 
   - If you have to do this, make sure this is documented explicitly or try to make them reversible


## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/bsGCgJr60as")`

<details>
<summary> Meeting chat log </summary>

```
00:45:32	collinberke:	Check this out regarding side effects: https://withr.r-lib.org/articles/changing-and-restoring-state.html
00:55:58	Arun Chavan:	https://rstats.wtf/
00:58:05	Isabella Velásquez:	https://yihui.org/knitr/options/
00:59:31	Arun Chavan:	+1 for knitr options because the tab autocompletion doesn’t work for them (at least for me)
01:00:58	Brendan Lam:	I'm gonna try that
```
</details>
