# Lifecycle

**Learning objectives:**

- How to determine package version and why it matters. 
- Difference between major and minor update
- Learn about breaking change and backward compatibility   
- How to communicate them to users

## Package evolution 

CRAN keeps history of package on GH and you can access it! 

### Package version number

- `package_version()` is a function that convert strings with either `.` or `-` to s3 object that deal with version numbers

```{r package_version} 
version <- package_version(c("1.0", c("0-9-1")))
class(version)
typeof(version)
```


- `utils::packageVersion()` allows us to get the version of a package: useful to do check on package in function 


- even if `-` is accepted it is not recommended 

- You should use at least 3 components on your numbering system 

## Backward compatibility and breaking change  


Let start with what is not a breaking change: 

- adding new functions, adding an arguments, fixing a bug!

If you do that do not worry (too much).

A breaking change will not allow reproducibility of previous codes with the new version. 

A good example is when you introduce a change and other package on CRAN that depebnd on it do not pass anymore the `R CMD check`.

Some examples are: 

- Removing a function 

- Removing an argument 

- Narrowing the set of valid inputs to a function 


## Major vs minor patch

- the numbering system help communicate change and evolution to users 

```
<major>.<minor>.<patch>        # released version
<major>.<minor>.<patch>.<dev>  # in-development version
```

- **Patch release**: small release, just bug fixing

- **Minor release**: the same + change that are backward compatible

- **Major release**: change that are not backward compatible, affect most users, stable API

You can use: `usethis::use_version()` to help you keep tracl on it 

## Pros and cons of breaking change

- useR usually do not care enought of the version of the package they use 

- `renv` is an option to keep track of it inside of project 

Deal with it! (and provide users fair warning and time see next part)


### Lifecycles, stages and badges

it is an other way to communicate to users. It introduces visual badges. 

It can be use at the package, function and arguments levels.

find more with `usethis::lifecycle()` function! 

### Deprecated a function

Lifecycle ha a vignette on it: [`vignette("communicate", package = "lifecycle")`](https://lifecycle.r-lib.org/articles/communicate.html)

You can generate warning and message to the users with `lifecyle` package: 

```
Warning message:
Problem while computing `..1 = across(where(is.factor) & !is.ordered, as.character)`.
ℹ️ Predicate functions must be wrapped in `where()`.

# Bad data %>% select(is.ordered)

# Good data %>% select(where(is.ordered))

ℹ️ Please update your code. This message is displayed once per session.
```

You should graduate the warning! 

### Dealing with change in a dependency 

A functionality you are using will be changed!!!!

One option: 

```{r eval=FALSE}
your_function <- function(...) {
  if (packageVersion("otherpkg") >= "1.3.9000") {
    otherpkg::their_new_function()
  } else {
    otherpkg::their_old_function()
  }
  # the rest of the function
}
```

### Superseding a function

Try to preserve old behavior but you want to let them know that they should change!!!


## SLIDE 1

- ADD SLIDES AS SECTIONS (`##`).
- TRY TO KEEP THEM RELATIVELY SLIDE-LIKE; THESE ARE NOTES, NOT THE BOOK ITSELF.

## Meeting Videos

This was a somewhat different chapter for cohorts 1-3.

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/Dh-kbknqa3E")`


### Cohort 2

`r knitr::include_url("https://www.youtube.com/embed/aU4pQTki_tg")`


### Cohort 3

`r knitr::include_url("https://www.youtube.com/embed/G852fyvKl8A")`

<details>
<summary> Meeting chat log </summary>

```
00:37:08	Rex Parsons:	https://fromthebottomoftheheap.net/2020/04/30/rendering-your-readme-with-github-actions/
00:59:25	Ryan Metcalf:	Was it `cast1?
```
</details>

### Cohort 3 Extra: Jenny Bryan

`r knitr::include_url("https://www.youtube.com/embed/5LktoXh7WvY")`

<details>
<summary> Meeting chat log </summary>

```
00:02:39	Jon Harmon (jonthegeek):	Questions: (link no longer available)
00:03:58	Ryan Metcalf:	I sent the link to Shel and Lucy too.
00:30:18	collinberke:	+1 for gargle
00:42:26	Rebecca Butler:	this was such a great question (and answer). thank you!
00:45:05	Ryan Metcalf:	https://github.com/search
00:50:32	Jon Harmon (jonthegeek):	https://r-universe.dev/search/
01:01:40	Jon Harmon (jonthegeek):	Vote up questions to help me choose the last one or two!
Questions: (link no longer available)
01:03:29	Jon Harmon (jonthegeek):	https://design.tidyverse.org/
01:04:10	Jon Harmon (jonthegeek):	https://style.tidyverse.org/error-messages.html
01:09:54	Jon Harmon (jonthegeek):	use_tidy_github_actions
01:10:34	Jon Harmon (jonthegeek):	use_tidy_upkeep_issue
01:13:42	Rebecca Butler:	Thank you so, so much!!
01:17:17	collinberke:	Thank you!
01:17:17	Brendan Lam:	Thank you!
```
</details>


### Cohort 4

`r knitr::include_url("https://www.youtube.com/embed/RIOdG96LyGc")`

<details>
<summary> Meeting chat log </summary>

```
00:55:28	Neil Birrell: usethis::use_lifecycle()
01:00:18	Neil Birrell: Warning message:
Problem while computing `..1 = across(where(is.factor) & !is.ordered, as.character)`.
ℹ️ Predicate functions must be wrapped in `where()`.

# Bad data %>% select(is.ordered)

# Good data %>% select(where(is.ordered))

ℹ️ Please update your code. This message is displayed once per session.
```
</details>
