--- 
title: "R Packages Book Club"
author: "The Data Science Learning Community"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: book.bib
biblio-style: apalike
link-citations: yes
github-repo: r4ds/bookclub-rpkgs
description: "This is the product of the Data Science Learning Community's R Packages Book Club."
---

# Welcome {-}

This is a companion for the book [_R Packages_](https://r-pkgs.org/) by Hadley Wickham and Jenny Bryan (2nd edition, copyright 2023).

-  These are **shared slides!**
   -  Developed by the [Data Science Learning Community](https://dslc.io).
   -  [Join the community](https://dslc.io/join) to participate.
   -  Available at [dslc.io/rpkgs](https://dslc.io/rpkgs).
-  We follow the [Data Science Learning Community Code of Conduct](https://dslc.io/conduct).

## Book club meetings {-}

-  [Volunteer leads discussion of a chapter](https://docs.google.com/spreadsheets/d/1C75auf7ceIt3zieY85Hmb1oVkVNtSzRoZTmSU7SVYRc/edit?usp=sharing)
   -  **This is the best way to learn the material.**
-  Presentations: 
   -  Review of material
   -  Questions you have
   -  Maybe live demo
-  More info about editing: [this github repo](https://github.com/r4ds/bookclub-rpkgs).
-  Recorded, available on the [Data Science Learning Community YouTube Channel](https://dslc.io/youtube).
-  **Camera** is **optional,** but **encouraged.**

## Pace {-}

-  Meet ***every*** week except holidays, etc
   -  We'll discuss even if presenter unavailable
-  Meetings = **1 hour.**
-  **Goal:** 1 chapter/week
-  Ok to split overwhelming chapters
-  Ok to combine short chapters
-  If we need to **slow down** and discuss, **let me know.**
   -  Most likely someone has the same question 
   -  We are all here to learn 

## Introductions {-}

If you feel comfortable sharing, unmute or raise your hand!

-  **Who** are you?
-  **Where** are you joining from?
-  **Previous clubs?** (DSLC or other)
-  **How long** have you been using R?
-  **What** are you most looking forward to learning?

## git and GitHub {-}

-  **Git/GitHub** (version control) = **valuable skill.**
   -  Discussed more in **Chapter 21.**
-  If you present, best to **integrate your materials** into this bookdown version.
-  You'll need a **GitHub account** to submit modifications.
-  **Full instructions** in [GitHub repository](https://github.com/r4ds/bookclub-rpkgs) ("repo").

## Group Question 1 {-}

(Submit answers in chat)

On a scale from 1 - 5, how familiar are you with using git for version control?

1 = Not at all familiar (a true beginner)

to

5 = Very familiar (use git everyday/can handle a merge conflict like a pro)

## Group Question 2 {-}

(Submit answers in chat)

On a scale from 1 - 5, how familiar are you with using GitHub?

1 = Not at all familiar (a true beginner)

to 

5 = Very familiar (I have GitHub account with repositories)

## Group Question 3 {-}

(Submit answers in chat)

On a scale from 1 - 5, how valuable would it be to devote a week to introducing git and GitHub?

1 = Not at all valuable

to 

5 = Very valuable

## git and GitHub Resources {-}

-  [Happy Git and GitHub for the useR](https://happygitwithr.com/)
-  [`usethis`'s pull request helpers](https://usethis.r-lib.org/articles/pr-functions.html)
-  [`git`'s documentation](https://git-scm.com/doc)
-  [MShiny Cohort 2 Introduction](https://www.youtube.com/watch?v=beOYuHG9Xng&list=PL3x6DOfs2NGjhwrYvdmrKRNcvXX7X6ldt)

## Learning objectives {-}

-  Students who study with LOs in mind ***retain more.***
-  **Tips:**
   -  "After today's session, you will be able to..."
   -  *Very* roughly **1 per section.**
-  Likely need to be refined

## Today's learning objectives {-}

-  Explain **what** an **R package** is.
-  Describe **why** we should **develop packages.**
-  Discuss the philosophy behind the use of the **`devtools` and `usethis` packages** in **package development.**

## What is an R package? {-}

-  R packages can be defined based on their utility:
   -  A means to bundle code, data, documentation, and tests that can easily be shared with others. 
   -  A package opens up the use of several workflow packages/tools.
   
-  R packages can be defined by their structure:
   -  Discussed throughout the book club
   -  At its core, packages are defined by the conventions of how files and directories are organized: `/R`; `DESCRIPTION` file; `/tests`; `/data`; etc. 
   -  At its core, packages are defined by the conventions of how files and directories are organized: `/R`; `DESCRIPTION` file; `/tests`; `/data`; etc. 

## Installing and using packages {-}

-  Packages are available via [Comprehensive R Archive Network (CRAN)](https://cran.r-project.org/index.html)
-  Packages are also available via different public repos
   -  [GitHub is the most popular](https://github.com/).

### Installing packages {-}

```{r 01-install-packages, eval=FALSE}
install.packages("x")
library("x")

# Getting help
package?x
help(package = "x")
```

## Why develop packages? {-}

-  Allows us to more easily share and use code amongst the team and other collaborators.
-  Saves yourself and your team's time.
-  Opens up the use of several tools to more efficiently write code.
-  Defines a formal organizational structure of where to put things. 
-  Allows us to develop more robust, rigorous, reproducible analyses.
-  Helps us write better code.

## Philosophy behind the tools of R Package development {-}

-  Anything that can be automated, should be automated. 
-  Use functions for package development, avoid doing it by hand. 
   -  To do this, we need to learn the functions/workflow. 
   -  Our thinking should be on what we want the package to do, not on how to organize a package. 
   -  This is where the [`devtools`](https://devtools.r-lib.org/) and [`usethis`](https://usethis.r-lib.org/) package come into play. 
   -  Tools to insulate us from the low-level details of how packages are built.
-  `devtools` works nicely with RStudio, so that's the development environment we will use. 

```{r 01-install-devtools-usethis, eval=FALSE}
install.packages(c("devtools", "usethis"))
```

## What if we need more detail? {-}

-  The official manual is [Writing R Extensions](https://rstudio.github.io/r-manuals/r-exts/).
   -  This document is exhaustive. 
   
## What we won't learn {-}

-  Important topics not covered in the book
   -  Git and Github
   -  Compiled code
   -  Markdown and Rmarkdown

## A note {-}

-  Don't get overwhelmed
   -  Some stuff you may never use.
   -  Remember, an R package at it's core is just an `R` directory.
  
-  You are not expected to know everything.
   -  We are all learning.
   -  None of us is as smart as all of us.
   -  It's okay to say "I don't know."
   -  We can stop at anytime to discuss.


## Meeting Videos {-}

### Cohort 1 {-}

`r knitr::include_url("https://www.youtube.com/embed/FR6NsbkYhcw")`


### Cohort 2 {-}

`r knitr::include_url("https://www.youtube.com/embed/7YgH8qfyzFU")`


### Cohort 3 {-}

`r knitr::include_url("https://www.youtube.com/embed/QOmdmGMxl_Q")`

<details>
<summary> Meeting chat log </summary>

```
00:07:46	Rex:	Brisbane, Australia
00:36:26	Collin Berke:	On a scale from 1 - 5, how familiar are you with using git for version control?
00:36:51	Ryan Metcalf:	4
00:36:52	Brendan Lam:	2
00:36:52	Arun Chavan:	3
00:36:53	Larissa Shaughnessy:	2
00:36:54	Isabella Velásquez:	4
00:36:57	Rex Parsons:	3
00:37:19	Collin Berke:	On a scale from 1 - 5, how familiar are you with using GitHub?
00:37:38	Rex Parsons:	4
00:37:38	Arun Chavan:	4
00:37:38	Isabella Velásquez:	5
00:37:39	Brendan Lam:	2
00:37:41	Larissa Shaughnessy:	3
00:38:14	Collin Berke:	On a scale from 1 - 5, how valuable would it be to devote a week to introducing git and GitHub?
00:38:29	Brendan Lam:	3
00:38:31	Arun Chavan:	2
00:38:35	Larissa Shaughnessy:	4
00:38:35	Ryan Metcalf:	4
00:38:36	Rex Parsons:	4
00:38:48	Isabella Velásquez:	I am comfortable with Git, but if it’s valuable for the group!
01:05:36	Rex Parsons:	Thanks, Collin!
```
</details>


### Cohort 4 {-}

`r knitr::include_url("https://www.youtube.com/embed/BaGzNVLqfAE")`

<details>
<summary> Meeting chat log </summary>

```
00:10:03	Schafer, Toryn:	https://docs.google.com/spreadsheets/d/161KcusuKQlNcmYyjVszV6hdzBSJzDA6GpulS_Bj5RwU/edit#gid=0
00:15:59	Schafer, Toryn:	Who are you?
Where are you calling in from?
How long have you been using R?
What are you most looking forward to during the group?
00:16:11	Jamie Hogg:	Hi all, I can't use my microphone right now. Apologies. I am a PhD student at QUT in Brisbane, Australia. I have been using R for about 2 years or so.
00:23:48	Schafer, Toryn:	2.5
00:23:50	Virgilio Gonzenbach:	Q1: 4 (dont handle merges like a pro :c )
00:23:51	Howard Baek:	3.3
00:23:54	Sofi GS:	2
00:23:55	defuneste:	3
00:23:59	Oluwafemi Oyedele:	3
00:24:04	Jamie Hogg:	maybe 3. Same here, never worked on a true collaborative project
00:24:15	Jamie Hogg:	2
00:24:20	Howard Baek:	5
00:24:23	Schafer, Toryn:	5
00:24:24	defuneste:	3
00:24:26	Sofi GS:	3
00:24:27	Virgilio Gonzenbach:	4
00:24:32	Oluwafemi Oyedele:	3
00:24:51	defuneste:	1
00:24:54	Virgilio Gonzenbach:	1.5
00:24:56	Schafer, Toryn:	2
00:24:58	Howard Baek:	1.4
00:24:59	Sofi GS:	2
00:25:06	Jamie Hogg:	3
00:25:21	Oluwafemi Oyedele:	4
00:32:50	Jamie Hogg:	where is this signup document your discussing?
00:33:03	Schafer, Toryn:	https://docs.google.com/spreadsheets/d/161KcusuKQlNcmYyjVszV6hdzBSJzDA6GpulS_Bj5RwU/edit#gid=0
00:33:35	Jamie Hogg:	great! Thank you
```
</details>


### Cohort 5 {-}

`r knitr::include_url("https://www.youtube.com/embed/iJahyHlPz0I")`

<details>
<summary> Meeting chat log </summary>

```
00:08:26	shamsuddeen:	https://docs.google.com/spreadsheets/d/1ygvVSCNuGRYy78KCtBsmmH0boZpWJi-XshTvIQMHe2s/edit
00:19:43	Trevin Flickinger:	GitLab and Bitbucket are two other public options for packages
00:27:29	Trevin Flickinger:	I’m okay with a two week break
00:27:44	Trevin Flickinger:	Advanced R is on a break those two weeks as well
```
</details>
